Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins

### 2017-06-06 Manuscript.tex
miscellaneous changes

\title[Multisequence Monte Carlo simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching}

@@

\keywords{Monte Carlo, generalized ensembles, simulated tempering, protein folding, protein evolution}

@@

\section{Introduction}
\noindent
Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} These advances have in part triggered major efforts~\cite{McGuffee2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the molecular dynamics of bio-systems of different sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} or a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future for detailed simulations.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation either of the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches which can further speed up conformational sampling, e.g., activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, a range of systems naturally call for the investigation and comparison of many molecular variants, e.g., determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} or protein-nucleotide interactions,~\cite{Rohs2010} or the role of mutations in disease processes such as protein aggregation.~\cite{Ross2004} Another example is protein folding where unique insight has been achieved by comparing the folding within and between protein families.~\cite{Tzul2017,Wensley2010} With the increasing availability of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

To this end, we evaluate in this work an MC-based algorithm that can in principle calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding. This ``multisequence" (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was later adapted for the characterization of peptide-protein binding specificity,~\cite{Bhattacherjee2013,Wallin2017}, but it has to our knowledge not been previously tested in protein folding simulations. The MS algorithm relies on sampling at fixed temperature a generalized ensemble in which the sequence is a dynamic parameter. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. As illustrated in Fig.~1, this approach is easily applied when $\rbar$ and $\sbar$ are ``perpendicular" coordinates such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$. While this may be hard to achieve for detailed protein models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model sequences with 16 amino acids. This set, {\SR}, was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and $\beta$-hairpin, respectively, and allowed us to demonstrate the existence of A1-N1 mutational pathways which abruptly switch between the two structures and do not proceed through unstable sequences.~\cite{Holzgrafe2014} Here we use this set to validate the MS method and compare its efficiency to a standard generalized-ensemble method, namely simulated tempering (ST).~\cite{Marinari1992,Lyubartsev1992} We thereafter apply the MS method to a greatly enlarged set with 1,024 sequences, also spanning the A1-N1 space, as well as another set spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively. Besides hinting at the feasibility of applying the MS method on large sets of sequences, the results allow us to carry out a more systematic analysis of the biophysical properties of sequences along mutational pathways connecting these two pairs of basic protein folds than has been possible with standard methods.~\cite{Holzgrafe2014,Holzgrafe2015}

%We find that sequences become gradually less stable as they approach switch point. However, fold switching is observed to be typically abrupt at the switch point. Folding pathways usually pass through bistable sequences which populate both folds up to a significant level. We show that while bistable sequences become less when selecting more stable pathways for A1-N1 system, they generally remain the same for the A2-TN system.

%We first apply the MS algorithm to a set of 144 model sequences with 16 amino acids and compare with the simulated tempering (ST) algorithm. We find that the computational efficiency of the MS algorithm is never less than ST except at the lowest studied temperature for one sequence.

%We then demonstrate that the multisequence algorithm can be used to calculate the thermodynamic behavior for $>1,000$ sequences in a 3-letter model for protein folding~\cite{Bhattacherjee2012} at temperatures well below the typical folding temperatures $T_\mathrm{f}$ of the included sequences.

%Modern investigations into biophysical properties of biomolecular systems are routinely carried out not only on a single species but on multiple molecular variants.~\cite{Vukmirovic2000,Nickson2010} For example, protein engineering methods are used to provide mechanistic insight into the specificity of protein-protein and protein-DNA interactions, to compare the  conformational propensities of disease-related mutations of aggregation-prone proteins, and optimize biomolecular properties through directed evolution. Moreover, large-scale screening techniques such as phage display and microarray chips are used to provide biophysical properties across entire genomes. In short, through the study of a biomolecular system and related variants molecular variants are studied in order to gain a more complete picture of the biophysical and biochemical properties of the system and provide evolutionary context.

%While protein or DNA sequences are easily changed on the computer, the high computational cost of simulating even a single moderately-sized biomolecule makes systematic explorations of sequence space challenging. Some notable efforts have nonetheless been attempted, including simulations of all 136 unique tetranucleotide sequences in B-DNA oligonucleotides using all-atom explicit water molecular dynamics.~\cite{Beveridge2004}

%While the 16-amino acid A1 and N1 sequences fold into $\alpha$-helical and $\beta$-hairpin folds, respectively, the two 35-amino acid sequences A2 and TN adopt the folds denoted IIA and IIB, as shown in Fig.~X.

@@

\subsection{Computational efficiency of the multisequence algorithm}
\noindent
We start by testing the MS algorithm on the set {\SR} consisting of the 144 3-letter model sequences with 16 amino acids studied in Ref.~\citenum{Holzgrafe2014}. Two of the sequences in {\SR} are A1 and N1 (see Table~I) which were designed to fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. A1 and N1 differ at 10 positions such that 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences were selected from this binary space with the constraint that the total number of h-amino acids are not too high and not too unevenly distributed.~\cite{Holzgrafe2014}

@@

We apply the multisequence algorithm on the set {\SR} across a range of temperatures (see Table~II). Figure~2 illustrates a typical MS simulation trajectory carried out at the lowest temperature which is well below the folding temperature of both A1 and N1~\cite{Holzgrafe2014,Holzgrafe2015}. From the MC evolution of the total energy $E$, sequence index $\sbar$, and RMSD values from the representative structures in Fig.~2B, it is evident that visits to various sequences drive transitions into a range of structural states. In particular, there are frequent visits to both $\alpha$-helix and $\beta$-hairpin structures. Transitions between them are accompanied by a shift in which sequences are preferably visited. For example, visits to high $\sbar$-indices, including N1 with index 144, tend to coincide with formation of $\beta$-hairpin structures.

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and high $\Delta h$, acceptance rates are only around 0.1-0.2 but are substantially higher at other $T$ and $\Delta h$ values. The overall acceptance rates are in fact above the oft-quoted rule-of-thumb value $0.25$~\cite{Gilks1996} for all temperatures $T$ (see Fig~4B). Increased acceptance rates can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer protein chains.

@@

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two sequences, R1 and R2, chosen randomly at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, a MS simulation provides the thermodynamics of all 144 sequences for one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). The MS algorithm is validated by comparing the average total energy, $\left < E \right >$, calculated for different $T$'s and $\sbar$'s, between the two sets of results (see Supplementary Information).

As a way to assess sampling efficiency, we compare in Figure~4 the statistical error, $\sigE$, of the average energy $\left <E\right >$ for the 4 different sequences obtained using ST and MS, respectively. Since roughly the same number of sampled conformations were obtained for each $\sbar$ and $T$ combination, we compare the statistical errors directly. The two algorithms give similar errors at the highest $T$ studied which implies consistent result between them. It should be noted that the highest $T$ studied is well above the folding temperature for all the 4 sequences. Therefore, this consistency is not particularly surprising since at high-T the free-energy landscape is relatively smooth. Hence, conformational space requires little difficulty to sample.

However, as temperature decreases, the statistical error from the MS algorithm generally becomes smaller than those from ST. In particular, Fig.~4C illustrates that MS is on average twice as accurate as ST in its error estimates for R1. The best estimate occurs at $\kb\approx$ 0.46 where MS is about 450\% better than ST. Even at the lowest T studied, MS is at least consistent or better than ST for all 4 sequences.

@@

%Hence, we conclude that in the comparison made above the computational efficiency of the MS algorithm is comparable to ST. It must be stressed, however, that while the ST simulation requires that the thermodynamic behavior is calculated at a range of temperatures, MS works well on its own directly at low temperatures. ST was introduced as an extension to the standard, fixed-$T$ Metropolis-Hastings algorithm in order to overcome the poor mixing of states at low temperatures, i.e., the tendency for the simulations at low $T$ to become trapped in local free energy minima. The MS algorithm, however, efficiently calculates the thermodynamics at low $T$'s without visits to other $T$'s. In the MS algorithm, it is visits to sequences with poor native state stability rather than high $T$'s that drive the decorrelation of generated states and escape from local traps. This way, calculation of low-$T$ thermodynamics is enhanced in MS simulations while at the same time enhancing coverage in sequence space. This makes the MS algorithm ideal for situations in which the finite-$T$ thermodynamic behavior of multiple sequences are of interest.

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence space}
\noindent
We now turn to the full binary sequence sets {\SI} and {\SII} with 1,024 sequences each. By applying the MS method to these two sets separately (see Table~II), we determined the thermodynamic behavior of each included sequence at relatively low temperatures. In particular, these simulations allowed us to find the stabilities relative to folds IA and IB ($\PIA$ and $\PIB$) for all sequences in {\SI} and the stabilities relative to folds IIA and IIB ($\PIIA$ and $\PIIB$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for A1-N1 and A2-TN, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and TN. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-TN, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-TN. We find that there are no possible A1-to-N1 or A2-to-TN pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

\subsection{Stability and bistability properties in of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural protein folding switching is that sequences near the switch point have reduced stability.~\cite{Bryan2010} The results from our model exhibit a similar trend. Fig.~6A and B show the total stability $\Ptot$ at different mutational distances between start and end points. Intermediate sequences are less stable than sequences at distances $h=0$ and $h=10$, although there are large variations as indicated by the spread between upper and lower bounds. Nonetheless, there is a clear statistical trend that sequences become gradually less stable as successive mutations are applied to any of the 4 start and end points until a minimum is reached. 
%These minima in stability occur closer to N1 and TN in the two systems, respectively, indicating that the IA/IIA folds are more mutationally robust than the IB/IIB folds.

@@

However, the smooth stability trends in Fig.~6A and B belie the real character of the individual mutational pathways which tend to exhibit an abrupt switch between the two folds. To examine the statistics of fold transitions in our model, we first make a distinction between two types of stable sequences: those that fold into a single unique fold, thus behaving as classical proteins, and those that exhibit significant stabilities of both folds. Such ``bistable" sequences are interesting from a biophysical perspective in that they alternate between populating two different folds and hence are highly dynamic proteins. Indeed, they have been proposed to play an important role in the evolution of new protein folds.~\cite{Sikosek2016}

We consider a stable sequence to be bistable if $B>0.5$, where $B$ is a bistability measure (see Fig.~6 legend). In principle, a fold transition can occur directly between two classical proteins with unique folds, or it can proceed via one or more intermediate bistable sequences which populate both folds. We  define the switch length of a mutational pathway $L_\mathrm{s}=2+n_\mathrm{B}$, where $n_\mathrm{B}$ is the number of bistable sequences in between the two classical sequences that define the switch point. Hence, a path with $L_\mathrm{s}=2$ accomplishes a fold switch in a single mutational step without going through a bistable point. From the distributions of  $L_\mathrm{s}$ in Fig.~5C and D taken over all pathways with $\Pcut=0.50$, it can be seen that fold switching along individual pathways are typically completed in only 1-2 mutations and often a single step is enough switch between the IA and IB folds. Hence, fold switching is typically abrupt and, for $\Pcut = 0.5$, it is fairly common that viable pathways pass through one or two bistable sequences.

Interestingly, IA-IB fold switching through one or more bistable sequences become less and less common as selections for more stable pathways are made. This can be seen from the decrease in $\left <L_\mathrm{s}\right >$ as a function of $\Pcut$ (see Fig.~5C(inset)). For $\Pcut\ge0.70$, there are no longer any viable pathways between the $\alpha$-helix and $\beta$-hairpin that passes through a bistable sequence because $\left <L_\mathrm{s}\right > =  2$. An underlying reason for the occurrence of sharper fold switches for more stable mutational pathways is apparent from a comparison between $\Ptot$ and $B$ across all sequences in {\SI}. As shown in Fig.~5E, sequences with the highest $\Ptot$ tend to exhibit very little bistability. Hence, highly stable mutational pathways are therefore forced to go through abrupt switch points, thereby transitioning between folds in a single mutational step. The situation for the IIA-IIB fold switch is more complicated. Sequences in {\SII} too follow the trend that high overall stability is achieved for only classical proteins with a single dominant fold. However, bistable sequences nonetheless play a crucial role in bridging the IIA-IIB folds because paths without bistable intermediate sequence are not available for $\Pcut\ge0.50$.  Hence, passing through sequences with bistable properties may still be required to achieve mutation-induced fold switching between certain fold pairs even through it lead to a reduced stability at the switch point.

@@

\section{Discussion}
\noindent
The simulation algorithm for biomolecules considered here works by making the sequence a dynamic parameter. We tested this approach by applying it to proteins. The results indicate that there are two main potential benefits. Firstly, it provides a convenient way to sample the respective canonical distributions for a large set of sequences in a single run and, secondly, it enhances the sampling of conformational space such that it can be applied directly at low temperatures.

The conformational sampling efficiency can be assessed from the comparison with ST. Although there is no single ``fair" way to compare the MS and ST methods, we chose as a measure of efficiency the statistical error of the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, which are well above the folding temperatures of both A1 and N1, the statistical errors $\sigE$ are basically the same for the two methods. This finding is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods leading to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant temperature. Hence, rather than promoting escape from local minima by visits to higher temperature, as is the case in  ST~\cite{Marinari1992,Lyubartsev1992} or temperature replica-exchange,~\cite{Swendsen1986}  MS simulations escape local minima through visits to sequences where the minima are not as deep or maybe altogether absent. For the same reason, we expect the performance of the MS algorithm to depend on the size the sequence set used, as well as their character.  Specifically, the MS simulation may hinge on the inclusion of at least a few sequences with poor stabilities such that partial unfolding of the chain is regularly triggered and thus promoting transitions to new conformational states.

>>>>

\title[Multisequence Monte Carlo simulations]{Multisequence algorithm for coarse-grained biomolecular simulations: exploring the sequence-structure relationship of proteins}

@@

\keywords{Monte Carlo, generalized ensembles, protein folding, protein fold switching}

@@

\section{Introduction}
\noindent
Recent years have seen important advances in biomolecular simulation methods, including improvements to standard molecular dynamics force fields,~\cite{Piana2014} the advent of several alternative atomistic simulation approaches,~\cite{Ding2008,Irback2006,Verma2009,Yang2007} and new techniques for  conformational sampling.~\cite{Bernardi2015} Together with the ever-increasing availability of computational resources, these advances have triggered a few major efforts~\cite{McGuffee2010,Miao2010,Perilla2016,Lindorff-Larsen2011,Yu2016} to characterize the dynamics of biomolecular systems of various sizes, e.g., a small native protein on the millisecond scale~\cite{Lindorff-Larsen2011} and a comprehensive model cytoplasm on the nanosecond scale.~\cite{Yu2016} While encouraging and insightful, these large-scale simulations have also highlighted the fact that severe tradeoffs in size and time scales will likely persist for the foreseeable future.

One way to expand the range of biomolecular simulations is to turn to coarse-grained (CG) models, where the basic aim is to simplify the physical description of interactions while retaining the essential physics of the system under study.~\cite{Riniker2012} Ingolfsson \textit{et al.} list 4 main factors that make CG models computationally fast: reduction in the number of degrees of freedom, faster simulation dynamics, emphasis on short-range interactions and the ability of using larger integration time steps.~\cite{Ingolfsson2014} To this list can be added that a CG representation of either the interaction potential or the molecular geometry often opens up for alternative sampling schemes beyond traditional molecular dynamics approaches, which can further speed up conformational sampling. Examples of such sampling schemes include activation-relaxation kinetics,~\cite{Beland2011} discrete molecular dynamics~\cite{Proctor2011} and various Monte Carlo (MC)-based techniques such as cluster moves.~\cite{Vitalis2009}

The challenges of achieving representative conformational sampling of individual biomolecular systems notwithstanding, many biological processes naturally call for the investigation and comparison of molecular variants. This is required, for example, in determining the molecular mechanisms of specificity in protein-protein~\cite{Zarrinpar2003,Hakes2007} or protein-nucleotide interactions,~\cite{Rohs2010} and the role of mutations in disease processes such as protein aggregation.~\cite{Ross2004} Another example is the protein folding process, where unique insight has been achieved by comparing the folding within and between protein families.~\cite{Tzul2017,Wensley2010} In a situation with rapid growth of sequence information,~\cite{Vukmirovic2000} it is of interest to explore ways to efficiently sample multiple sequences in biomolecular simulations.

To this end, we consider in this work an MC-based algorithm that can calculate the thermodynamics of multiple sequences in a single run and apply it to a coarse-grained model for protein folding. This multisequence (MS) method was originally developed in the context of homo- and heteropolymer simulations~\cite{Irback1995} and was also later adapted for the characterization of peptide-protein binding specificity.~\cite{Bhattacherjee2013,Wallin2017} To our knowledge, however, it has not been previously tested in realistic protein folding simulations. The MS algorithm carries out a simulation in a generalized ensemble that performs a random walk in sequence space. Hence, there are two main types of updates: conformational updates $\rbar\rightarrow\rbar'$ and sequence updates $\sbar\rightarrow\sbar'$. As illustrated in Fig.~1, this strategy can be applied when $\rbar$ and $\sbar$ are ``perpendicular" coordinates such that the potential energy of the model can be written in terms of two independent variables, $E(\sbar,\rbar)$. While this may be hard to achieve for detailed biomolecular models, it is likely possible in many CG models.

As an initial test case, we apply the MS method to a set of 144 model protein sequences with 16 amino acids. This set (denoted here {\SR}) was constructed to sparsely cover the sequence space between two ideally designed sequences, A1 and N1, which fold into an $\alpha$-helix and a $\beta$-hairpin, respectively.~\cite{Bhattacherjee2012} In a previous work, this set allowed us to demonstrate the existence of A1-N1 mutational pathways which do not proceed through disordered (non-folding) sequences but instead switch abruptly between the two folds in a single mutational step.~\cite{Holzgrafe2014} Moreover, it allowed the exploration of the biophysical factors that underpin this ``fold switching" phenomenon,~\cite{Holzgrafe2015} which was recently demonstrated in a handful of natural and engineered proteins.~\cite{Bryan2010} Here we use the set {\SR} and our CG model for protein folding to validate the MS method and compare its efficiency to a standard generalized-ensemble method.~\cite{Marinari1992,Lyubartsev1992} We thereafter greatly enlarge {\SR} to a set with 1,024 sequences, also spanning the A1-N1 space, as well as another set of the same size spanning two 35-amino acid sequences, A2 and TN, that fold into two-helical bundle and mixed $\alpha$-$\beta$ structures, respectively. Besides showing that the MS method can be applied to large numbers of sequences, the results allow us to carry out a more systematic analysis of the biophysical character of sequences along mutational pathways connecting these two pairs of basic folds than has been previously possible.

@@

\subsection{Computational efficiency}
\noindent
We start by applying the MS algorithm to the set {\SR} across a range of temperatures $T$ (see Table~II). Two of the sequences in {\SR} are A1 and N1 (see Table~I) which fold into stable $\alpha$-helix and $\beta$-hairpin structures, respectively, as shown in Fig.~2B. A1 and N1 differ at 10 positions such that 10 consecutive point mutations can transform A1 into N1, and vice versa. The binary sequence space between A1 and N1 in which any combination of these mutations have been carried out, therefore contains $2^{10}=1,024$ sequences. The 144 sequences in {\SR} were selected from this binary space with the constraints that the total number of hydrophobic amino acids are not too high and that they are not too unevenly distributed.~\cite{Holzgrafe2014}

@@

Figure~2 illustrates a typical MS simulation trajectory carried out at the lowest studied temperature which is below the folding temperature of both A1 and N1.~\cite{Holzgrafe2014,Holzgrafe2015} From the MC evolution of the total energy $E$, sequence index $\sbar$, and RMSD values from the representative structures in Fig.~2B, it is evident that visits to various sequences drive transitions into a range of structural states. In particular, there are frequent visits to both $\alpha$-helix and $\beta$-hairpin structures and transitions between them are accompanied by a shift in which sequences are preferably visited. For example, visits to high $\sbar$-indices, including N1 with index 144, tend to coincide with formation of $\beta$-hairpin structures as required to generate the correct equilibrium conformational ensembles.

One might have suspected that the MS algorithm would be hampered by poor acceptance rates for sequence updates. However, this is not the case in our model. We carry out updates $\sbar\rightarrow\sbar'$ by picking a new random sequence $\sbar'\ne\sbar$ from the set of allowed sequences. The (average) acceptance rate  depends on both $T$ and the step in sequences space $\Delta h$, i.e., the number of amino acid positions changed, as shown in Fig.~3. At the lowest $T$ and highest $\Delta h$, acceptance rates are only around 0.1-0.2. However, for most other $T$ and $\Delta h$ the overall acceptance rate is substantially higher and often above the oft-quoted rule-of-thumb value 0.25~\cite{Gilks1996} (see Fig~4B). We note therefore that increased acceptance rates can easily be achieved by restricting proposed updates such that $\Delta h\le\Delta h_\mathrm{max}$, where $\Delta h_\mathrm{max}$ is a maximum step size, which might be necessary for longer chains.

@@

We now compare the results from our MS calculations with simulated tempering (ST) simulations carried out on 4 of the 144 sequences, namely A1 and N1 and two random sequences, R1 and R2, chosen at distances $h=4$ and $h=6$ from A1, respectively (see Table~I). While ST provides the thermodynamics of a given sequence across a range of $T$ in a single run, an MS simulation provides the thermodynamics of all 144 sequences at one $T$. We adjust the simulation lengths for ST and MS runs such that roughly the same number of sampled conformations are obtained for each $\sbar$ and $T$ combination, thus ensuring that similar computational resources are used for the two algorithms (see Table~II). We first validate the MS algorithm by comparing the average total energy, $\left < E \right >$, calculated for these 4 sequences with the two different methods (see Supplementary Information). The two sets of results are entirely consistent showing that, for a given $\sbar$ and $T$, both MS and ST indeed sample the same distribution.

As a way to assess conformational sampling efficiency, we compare in Fig.~4 the statistical error, $\sigE$, of the average energy $\left <E\right >$ for the 4 sequences obtained using ST and MS, respectively. Because approximately the same number of sampled conformations were obtained for each combination of $\sbar$ and $T$, we compare the statistical errors directly. At the highest studied $T$, which is well above the folding temperature of both A1 and N1, the two algorithms give almost identical statistical errors. This can be understood by noting that at high-$T$ the free-energy landscape is relatively smooth and conformational space requires little difficulty to sample. The benefit of adding a dynamic parameter, whether $\sbar$ or $T$, is apparently minor under these conditions. However, at lower $T$, the $\sigE$ values from MS is often smaller than those from ST and never significantly higher. For example, at the lowest $T$, the precision in the estimate of $\left < E\right >$ is roughly twice as high in MS than ST for A1 and R1, and roughly the same for N1 and R2.

%The best estimate occurs at $\kb\approx$ 0.46 where MS is about 450\% better than ST. Even at the lowest T studied, MS is at least consistent or better than ST for all 4 sequences.

@@

\vspace{12pt}
\subsection{Exploring sequence space: IA/IB and IIA/IIB fold connectivities}
\noindent
We now turn to the full binary sequence sets {\SI} and {\SII} with 1,024 sequences each. By applying the MS method to these two sets (see Table~II), we determine the low-$T$ thermodynamic behavior of each included sequence. In particular, we calculate the stabilities of folds IA and IB, $\PIA$ and $\PIB$, for all sequences in {\SI} and the stabilities of folds IIA and IIB, $\PIIA$ and $\PIIB$, for all sequences in {\SII} (see Methods). The relative statistical errors on these quantities vary but are only a few percent at the most, despite the large number of sequences included.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to abrupt IA-IB or IIA-IIB fold changes, i.e., paths that do not pass through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at a single amino acid position. To determine if a sequence is stable we use the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for the IA-IB and IIA-IIB fold pairs, respectively; $\Ptot$ thus indicates the total stability of a sequence across the two competing folds. The precise network depends, of course, on the cut-off value of $\Pcut$ and a higher $\Pcut$ generally means a selection of more stable pathways.

Fig.~5 illustrates the networks obtained with $\Pcut=0.50$ showing that, at this stability threshold, there are mutational pathways that connect both IA-IB and IIA-IIB. A precise analysis shows that there are 516,972 viable IA-IB paths and 57,912 viable IIA-IIB paths. Because there are $10! =3,628,800$ possible paths between start and end points in both cases, they represent 14.2 \% and 1.6 \% of all possible paths of the A1-N1 and A2-TN sequence pairs, respectively. Hence, folds IA and IB are rather highly connected in our model for $\Pcut= 0.50$. For $\Pcut=0.60$, the numbers are 104,640 paths (2.9\%) for IA-IB and 22,512 (0.6\%) paths for IIA-IIB. We find that there are no possible IA-to-IB or IIA-to-IIB paths when $\Pcut\ge0.74$ and $\ge 0.66$, respectively. \\

\subsection{Biophysical properties of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural proteins that exhibit mutation-induced fold switching is a reduced stability near the switch point.~\cite{Alexander2009,He2012,Kouza2012,Sikosek2016,Sutto2012} Our model proteins exhibit a similar trend. Fig.~6A and B show the average total stability $\Ptot$ for sequences found at different Hamming distances $h$ from the starting point. Intermediate sequences are less stable than sequences at distances $h=0$ (A1 or A2) and $h=10$ (N1 or TN), although there are large variations between sequences as indicated by the spread between upper and lower bounds. There is nonetheless a clear statistical trend that sequences become gradually less stable as successive mutations are applied to any of the 4 start and end points until a minimum is reached. 
%These minima in stability occur closer to N1 and TN in the two systems, respectively, indicating that the IA/IIA folds are more mutationally robust than the IB/IIB folds.

@@

However, the smooth stability trends in Fig.~6A and B belie the real character of the individual mutational pathways which tend to exhibit an abrupt switch between the two folds. To see this and to further examine the character of the fold transitions in our model, we make a distinction between two types of stable sequences: those that fold into a single unique fold, thus behaving as classical proteins, and those that display substantial stabilities of both folds. Such ``bistable" sequences are interesting from a biophysical perspective in that they might continuously alternate between two different folds in a highly dynamic and coordinated fashion. Indeed, bistable sequences have been proposed to play a role in the evolution of new protein folds.~\cite{Sikosek2012} We consider a stable sequence to be bistable if $B>0.5$, where $B$ is a bistability measure (see Fig.~6 legend). In principle, a fold transition can occur directly between two classical proteins with unique folds, or it can proceed via one or more intermediate bistable sequences which populate both folds. We define the switch length of a mutational pathway $L_\mathrm{s}=2+n_\mathrm{B}$, where $n_\mathrm{B}$ is the number of bistable sequences in between the two classical sequences that define the switch point. Hence, a path with $L_\mathrm{s}=2$ accomplishes a fold switch in a single mutational step without going through a bistable point. From the distributions of  $L_\mathrm{s}$ in Fig.~5C and D, taken over all pathways with $\Pcut=0.50$, it can be seen that fold switching along individual pathways are typically completed in only 1-2 mutations and a single step is often sufficient to switch between the IA and IB folds. Hence, fold switching is typically abrupt and, for $\Pcut = 0.5$, it is fairly common that viable pathways pass through one or two bistable sequences.

Interestingly, switching between folds IA and IB through one or more bistable sequences become less and less frequent as selections for more stable pathways are made. This can be seen from the decrease in $\left <L_\mathrm{s}\right >$ as a function of $\Pcut$ (see Fig.~5C(inset)). For $\Pcut\ge0.70$, there are no longer any remaining path between the $\alpha$-helix and $\beta$-hairpin that passes through a bistable sequence because $\left <L_\mathrm{s}\right > =  2$. An underlying reason for the occurrence of sharper fold switches for more stable mutational pathways is apparent from a comparison between $\Ptot$ and $B$ across all sequences in {\SI}. As shown in Fig.~5E, sequences with the highest $\Ptot$ tend to exhibit very little bistability. Hence, highly stable paths are therefore forced to go through abrupt switch points where they transition directly between folds in a single step. The situation for the IIA-IIB fold pair is more complicated. We find that, just as for {\SI}, sequences in {\SII} follow the trend that the highest $\Ptot$ values occur for only classical, low-$B$ proteins. One might therefore expect that selection of more stable IIA-IIB paths would decrease $\left <L_\mathrm{s}\right >$, however, this is not the case as such abrupt switch points between the IIA and IIB are not available for $\Pcut\ge0.50$ (cf. Fig.~5 bottom). As a result, bistable sequences do play a crucial role in bridging the IIA and IIB folds, although passing though these sequences lead to additional reduction in stability at the switch point.

%. Hence, passing through sequences displaying substantial bistability may still be required to mutationally drive switching between certain fold pairs as single-step transitions are not available, thereby

@@

\section{Discussion}
\noindent
We have evaluated a biomolecular simulation algorithm that works by making the sequence a dynamic parameter by testing it on a CG model for protein folding. The results indicate that there are two main potential benefits of this approach. Firstly, it provides a convenient way to sample the canonical distributions of a large set of sequences in a single run and, secondly, it enhances the sampling of conformational space such that it can be applied directly at low temperatures. The conformational sampling efficiency can be assessed from the comparison with ST. Although there is no single ``fair" way to compare the MS and ST methods, we chose as a measure of efficiency the statistical error of the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, we find that the statistical errors $\sigE$ are basically the same for the two methods. This finding is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods leading to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant $T$. Hence, rather than promoting escape from local minima by visits to higher $T$, as is the case in  ST~\cite{Marinari1992,Lyubartsev1992} or temperature replica-exchange,~\cite{Swendsen1986}  MS simulations escape local minima through visits to sequences where the minima are not as deep or may be altogether absent. For the same reason, we expect the performance of the MS algorithm to depend on the size the sequence set used, as well as their conformational properties.  Specifically, the performance of MS simulations may hinge on the inclusion of at least a few sequences with poor stabilities such that partial unfolding of the chain is regularly triggered and thus promoting transitions to new conformational states.
